DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2017-01-12 2015CompleteDerivation/completeDerivation.tex
Another pass at the intro with Christine

Status:
\begin{itemize}
\item Section \ref{sec:fundamental_model}: Done.

\item Section \ref{sec:dynamical_systems}: Done.

\item Section \ref{sec:reducibility}: Done.

\item Section \ref{sec:relativity}: Done.

\item Section \ref{sec:Lagrangian}: Mostly done. Needs review.

\item Section \ref{sec:quantum_prelude}: Mostly done. Needs review.

\item Section \ref{sec:irreducibility}: Mostly done. Needs review.
\end{itemize}

%While some branches of physics (e.g. Newtonian mechanics, thermodynamics, special relativity) are derived from physical principles (e.g. Newton's laws, laws of thermodynamics, invariance of the speed of light), others start by simply setting a mathematical framework. This is the case for classical Lagrangian mechanics, where we start by assuming the existence of a Lagrangian function whose path integral is minimized during the motion, for classical Hamiltonian mechanics, where we start by assuming the state is described by conjugate pairs $(q^i, p_i)$ and evolves according to Hamilton's equations, and for quantum mechanics, where the state space is a complex inner product space and the state evolves according to Schroedinger's equation. That is, at the very core of some of our physical theories we have mathematical definitions instead of physical ideas.

%If we ask ``when can a particular physical system be suitably described by classical Hamiltonian mechanics?" the only rigorous answer we have is ``when it fits the mathematical framework". While many seem to be content with this answer (e.g. ``you have to start from somewhere, it may as well be the math which at least is precise"), it begs the question: it does not tell us how and what physical properties of the system under study are being captured.

The past century has seen a resounding success in the use of mathematics in fundamental physics. In many cases, it was mathematical ideas that allowed progress in theoretical physics that ultimately led to experimental discoveries. This success should, and indeed is, celebrated but, however, we should keep in mind that it comes at a price.

The issue is that mathematical structures are not enough to characterize what physical system is being studied. For example, the same mathematical framework for linear circuits can be used to describe electric, hydraulic, thermal or mechanical systems~\cite{Roland,Barron,Borutzky}. The mathematical treatment is the same because it captures aspects of the description that are common to all cases, that indeed is its power. But for the very same reason, we cannot infer what system is being described if all we are given is the mathematical model. In the jargon of linguistics/computer science/philosophy~\cite{Freidin,Friedman}: math captures the syntax (the relationships between the objects) but does not capture the semantics (the meaning of each object).

@@

This means that if we have a system in front of us, we do not have a rigorous way to conceptually establish whether the system is Hamiltonian or not, whether it is classical or quantum. We have some heuristics we can apply, but, at the end the day, a system is Hamiltonian because it follows the Hamiltonian framework.\footnote{The lack of a properly defined, and uniquely accepted, semantic is of course very evident in quantum mechanics with its numerous interpretations.}

Unsatisfied with this situation, we started creating a sort of math-to-physics dictionary so that each mathematical concept could be clearly associated to a crisp physical concept. As we pushed furthered back toward the basic mathematical definitions, we ended up with physical concepts of a more general nature and found that, sometimes, a single physical idea could have more than one mathematical consequence. That is: by putting physics at the center, we could achieve a sort of unification of ideas that was not evident in the math alone. The unexpected success of the approach led us to think that, maybe, the picture we were developing was a better framework to understand and link different parts of basic science and mathematics. This work is the systematized result of this endeavor.

%We started this work with the unpretentious goal of exploring simple questions such as the following. What do the mathematical symbols used in classical/quantum Hamiltonian mechanics represent? What is the physical meaning attached to them? Why is the state of a quantum particle described by a wave function (i.e. a vector in a complex inner product space), while the state of a classical particle is described by position and conjugate momentum (i.e. a point in a cotangent bundle)? The idea was to create a math-to-physics dictionary so that each mathematical concept was clearly mapped to a crisp physical concept. But to make sure that such a dictionary was complete, we realized we had to make sure that the whole relevant mathematical framework could be recovered: the translation needs to work both ways.

%This redefined our mission to rederiving as much as possible of the mathematical structure by starting from the smallest set of physical assumptions. It also forced us to push further and further back toward the basic definitions. For example, if we truly wanted to justify why the state space of Hamiltonian mechanics is charted by conjugate variables $q^i$ and $p_i$, we had to justify why it was a manifold as well, which in turn meant we needed to justify why it was a topological space. But when we did find physical concepts that mapped to these more basic mathematical definitions, we realized that this link was valid for a far wider range of cases: the concepts were more general than we anticipated. For example, we realized that the mathematical notion of a topology maps incredibly well to the idea of physical distinguishability, but states of Hamiltonian system are not the only things that are physically distinguishable.

%Each new link often shrank the complexity of the arguments while expanding the scope of the work in a way that was not at all anticipated. The same link that allowed us to justify Hamiltonian dynamics also gave us relativistic motion on the extended phase space. The same link that allowed us to justify Lagrangian mechanics also constrained the motion to that of massive particles under scalar/vector potential forces. We also realized that sometimes different mathematical concepts could be justified from the same physical starting point. That is: by putting physics first we could achieve a sort of unification of ideas that was not evident in the math alone.

%The unexpected success of the approach led us to think that, maybe, the picture we were developing was not simply ``a nicer way to think about things", but instead a better framework to understand and link different parts of basic science and mathematics. After many years of work and different incremental iterations, we think the arguments are well developed enough to report the progress.

The main goal is to show that it is possible to derive classical/quantum Hamiltonian/Lagrangian particle mechanics starting from few physical assumptions. We do \emph{not} claim that we have perfectly achieved this result, with no room for improvement. We \emph{do} claim that the whole architectural framework is solid enough to be convinced that not only this is indeed possible, but it allows a more unified picture that facilitates intuitive connections between different areas of physics and mathematics. It enables a consistent narrative across several domains of knowledge.

Rederiving classical, quantum and (to some extent) relativistic mechanics within the same work is therefore crucial to show that the concepts we develop are truly general, that they can be applied to many different cases. If we could derive classical Hamiltonian mechanics on assumptions that are inapplicable to the quantum case, what good would it do? Quantum systems are Hamiltonian systems, so clearly we wouldn't have understood what makes a system Hamiltonian. The breadth of applicability is what makes the work valuable.

>>>>

Status:
\begin{itemize}
%\item Section \ref{sec:fundamental_model}: Done.

%\item Section \ref{sec:dynamical_systems}: Done.

%\item Section \ref{sec:reducibility}: Done.

%\item Section \ref{sec:relativity}: Done.

\item Section \ref{sec:Lagrangian}: Needs review.

\item Section \ref{sec:quantum_prelude}: Needs review.

\item Section \ref{sec:irreducibility}: Needs review.
\end{itemize}

The past century has seen a resounding success in the use of mathematics in fundamental physics. In many cases, mathematical ideas were what allowed progress in theoretical physics and ultimately led to experimental discoveries. While these ideas are indeed useful---anything that can advance our understanding is welcome---they cannot, however, be the whole story.

The issue is that mathematical structures are not enough to characterize what physical system is being studied. For example, the same mathematical framework for linear circuits can be used to describe electric, hydraulic, thermal or mechanical systems~\cite{Roland,Barron,Borutzky}. The mathematical treatment is the same because it captures aspects of the description that are common to all cases; that indeed is its power. But for the very same reason, we cannot infer what system is being described if all we are given is the mathematical model. In the jargon of linguistics/computer science/philosophy~\cite{Freidin,Friedman}: math captures the syntax (the relationships between the objects) but does not capture the semantics (the meaning of each object).

@@

This means that if we have a system in front of us, we do not have a rigorous way to conceptually establish whether the system is Hamiltonian or not, whether it is classical or quantum. We have some heuristics we can apply, but, at the end the day, a system is Hamiltonian because it follows the Hamiltonian framework.\footnote{The lack of a properly defined, and uniquely accepted, semantics is of course very evident in quantum mechanics with its numerous interpretations.}

Unsatisfied with this situation, we started creating a sort of math-to-physics dictionary so that each mathematical concept could be clearly associated to a crisp physical concept. As we pushed further back toward the basic mathematical definitions, we ended up with physical concepts of a more general nature and found that, in some cases, a single physical idea could have more than one mathematical consequence. That is: by putting physics at the center, we could achieve a unification of ideas that was not evident in the math alone. The unexpected success of the approach led us to think that, perhaps, the picture we were developing was a better framework to understand and link different areas of basic science and mathematics. This work is the systematized result of this endeavor.

The main goal is to show that it is possible to derive classical/quantum Hamiltonian/Lagrangian particle mechanics starting from few physical assumptions. We do \emph{not} claim that we have perfectly achieved this result, with no room for improvement. We \emph{do} claim that the whole architectural framework is solid enough to be convinced that not only is this indeed possible, but it allows a more unified picture that facilitates intuitive connections within and across different areas of physics and mathematics. It enables a consistent narrative across several domains of knowledge.

Rederiving classical, quantum and (to some extent) relativistic mechanics within the same work is therefore crucial to show that the concepts we develop are truly general, that they can be applied to many different cases. If we could derive classical Hamiltonian mechanics on assumptions that are inapplicable to the quantum case, what good would it do? Quantum systems are Hamiltonian systems, so clearly we wouldn't have understood what makes a system Hamiltonian. This breadth of applicability is what makes the work valuable.
